\large How resource abundance and stochasticity affect animals' spatial needs

### 2023-07-05 writing/manuscript.Rmd
completed manuscript

---
title: \large How resource abundance and stochasticity affect animals' spatial needs
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: ubco
  - name: Adam T. Ford
    institute: ubco
  - name: Jacob R. Goheen
    institute: wyom
  - name: E. Patrícia Medici
    institute: [tapir, iucn, escas]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: ubco
institute:
  - ubco: The Irving K. Barber Faculty of Science, The University of British Columbia, Okanagan Campus, Kelowna, Canada.
  - wyom: Department of Zoology and Physiology, University of Wyoming, Laramie, Wyoming, United States of America
  - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
  - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
  - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
bibliography: 'env-var-review.bib'
csl: 'the-american-naturalist.csl'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    # - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{wrapfig} # for wrapping text around figures
    - \usepackage{caption} # for more customization of captions
    # smaller caption text, less space above and below caption
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{afterpage} # for \afterpage{\<command>}, e.g. \afterpage{\clearpage}
    # - \usepackage{gensymb} # for \degree, \celsius, \perthousand, \micro and \ohm
    # - \usepackage{float} # for better floating objects interface; adds more floats     # - \usepackage{ragged2e} # for \Centering, \RaggedLeft, and \RaggedRight
subparagraph: true # needed for \usepackage[small]{titlesec}
linenumbers: true
output:
  bookdown::pdf_document2:
    latex_engine: xelatex # @reed_frequency_2003 has unicode character U+0308
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: false
---

@@

\newcommand*\vpos{\var (\text{position})} <!-- positional variance -->

@@

<!--
@ciuti and @berger-tal: human presence in nature for reasons other than hunting
"When habitats ensuring different functions (Dunning et al., 1992; Camp et al., 2013) are far from each other..."
-->

@@

\noindent Animals' spatial needs are thought to be tightly linked to the amount of resources within their habitats, such that animals living in productive habitats generally require less space than those in resource-poor habitats. Although this hypothesis has widespread empirical support, existing studies have focused primarily on responses to the mean amount of resources, while responses to the variance around the mean are still largely unknown. This is not a trivial oversight. Animals adjust to variable environmental conditions, so failing to account for the effect of resource unpredictability can result in a limited understanding of animals' spatial needs. In this study, we first review the literature of studies that estimate animals' spatial needs as a function of resource abundance. We then leverage this information to provide a unifying framework and hypotheses for the effect of mean *and* variance in resources on animals' space use. Next, we use simulated movement data to demonstrate how the combined effects of mean and variance in resource abundance interact to shape predictable patterns in animal space use. Finally, we use real-world tracking data on a lowland tapir (\textit{Tapirus terrestris}) from the Brazilian Cerrado to show how this framework can be applied to better understand the movement ecology of free-ranging animals. Results from the simulations and empirical examples are presented using a fully transparent approach that allows researchers to apply the framework to their own data and inform area-based conservation efforts.

@@

\noindent The amount of resources an organism is able to access is a strong determinant of its odds of survival and reproduction. Resource limitations can cause individuals to experience a negative energetic balance, which can then result in lower fitness [@hou_cold_2020; @le_bot_fishery_2019], changes to community structure [@haney_tipping_2018; @burson_competition_2018; @riotte-lambert_environmental_2020; @ghislandi_resource_2018], altered physiology [@dai_pra_ground_2022; @rocha_life_2021; @le_bot_fishery_2019; @wessling_seasonal_2018], lower chance of reproduction [@stefanescu_timing_2021; @schmidt_interplay_2020; @le_bot_fishery_2019; @douglas_relative_2014], and even death [@foley_severe_2008; @berger_climate_2018]. Thus, many organisms will adapt their behaviors in response to changes in local resource abundance. Some species may respond to fluctuations in resource abundance by, if possible, switching to other food sources [@steinmetz_effects_2021; @le_bot_fishery_2019], and reducing energetic costs by reducing body heat [@schmidt_interplay_2020] or entering hibernation or torpor [@fjelldal_nightly_2021; @mohr_cellular_2020; @boyles_optimal_2020], but movement represents one of the most readily available traits that species can adjust, whether this be by modifying their home range [@bista_effect_2022; @bradsworth_using_2022; @lucherini_habitat_1996; @relyea_home_2000; @yu_artificial_2022; @arechavala-lopez_common_2019], migrating [@middleton_green-wave_2018; @geremia_migrating_2019], dispersing [framework: @southwood_habitat_1977; amphibians: @cayuela_determinants_2020; birds: @pretorius_movement_2020; @ref; mammals: @ref], or moving nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019].

While there are many other causes of animal movement, the relationship between animal movement and resource abundance has been of interest to biologists for over half a century. @burt_territoriality_1943 considered the search for food as the primary driver for animal movement within an animal's home range. Three decades after, @southwood_habitat_1977 suggested change in resource abundance is a strong determinant of how animals decide when and where to live and reproduce. Two years later, @harestad_home_1979 proposed that the simplest relationship between resource abundance and an animal's home range size is

\begin{equation} \label{harestad-eq}
H = C / R,
\end{equation}

\noindent where $H$ is the animal's **home range size**, $C$ is the animal's resource consumption, and $R$ is the abundance of (usable) resources. More recently, various researches reported that home range size decreases linearly with resource abundance [desert mule deer: @relyea_home_2000; red panda: @bista_effect_2022]. Others demonstrated that spatial needs decrease nonlinearly as habitat productivity increases, with stronger decreases when resources are scarce [e.g., Canadian lynx: @herfindal_prey_2005; carnivores: @nilsen_can_2005; tigers: @simcharoen_female_2014; ferruginous hawks: @watson_ferruginous_2020]. Additionally, an animal's spatial requirements depend strongly on its metabolic rate, diet, and body weight [@harvey_primate_1981; @lindstedt_home_1986; @reiss_scaling_1988; @harvey_primate_1981; @gittleman_carnivore_1982; @boratynski_energetic_2020; @noonan_effects_2020]. Thus, there are many drivers of animal movement and many causes of variety between behaviors, and we currently lack a unifying framework for quantifying the effects of resource abundance and unpredictability on animals' spatial needs. While much work has been done on estimating animals' responses to resource abundance, there is little to no information on they respond to unpredictable changes in resources.

In this paper, we refer to a location's average amount of available and accessible resources as **resource abundance**, while use the phrase **resource unpredictability** to indicate the variability in resource abundance due to *unpredictable* causes. We argue that, on its own, a habitat's long-term resource abundance is not sufficient to assess the habitat's quality. For instance, a grassland with relatively low but constant forage availability will clearly require drastically different behaviors and adaptations from a desert location with rare, sudden, and strong pulses of resources. In the first scenario, an animal may require a large but constant home range as it moves between locations in search of food, while an animal in the second scenario may switch between dispersal as it searches for high-resource patches and short-term range residency until the local resources are exhausted. Although it may be possible for both habitats to have the same long-term average resource abundance, the differences in resource unpredictability result in substantially different movement strategies being selected for. Although it is generally expected that resource unpredictability will decrease animals' fitness and a landscape's energetic balance [@chevin_adaptation_2010], there is little empirical evidence to support this hypothesis [but see: @herfindal_prey_2005; @nilsen_can_2005; @rizzuto_forage_2021].

In the following sections, we illustrate how animals' spatial needs depend on the abundance and unpredictability of resources. First, we set the theoretical background necessary for the successive sections, and concepts and notation are introduced without assuming any previous knowledge of statistical notation or movement ecology. Next, we provide a review of the effects of resource abundance on animals spatial needs while suggesting a unifying hypothesis. Afterwards, we provide a review of the effects of resource unpredictability on animals spatial needs while suggesting a second unifying hypothesis. Subsequently, we test the two hypotheses using quantitative simulated animal responses to changes in resource abundance and unpredictability. Finally, we demonstrate how this framework can be used in practice to describe the movement ecology of a lowland tapir living in the Brazilian Cerrado. Results from the simulations and empirical example are presented using a fully transparent approach that allows researchers to replicate the work and apply the methods to their own tracking data.

@@

In this section, we presents the foundational concepts and terms used in this manuscript. We introduce each concept without assuming a strong statistical or quantitative background to maximize accessibility and minimize misunderstandings and misinterpretations.

@@

\noindent In statistics, **random variables** indicate random (i.e., unknown) quantities and are indicated with capital letters (e.g., $R$). Known values, such as realizations of random variables (i.e., known observations or instances), are indicated with lower-case letters (e.g., $r$). Using this notation, we can write the statement "the probability of random variable $R$ taking the value $r$" as $P(R = r)$. Since resource abundance depends on a variety of factors, including climate, weather, competitive pressure, and individual- and species-specific energetics, it is often unpredictable (and hard to quantify). Thus, we can let the random variable $R$ indicate the *resource* abundance at a given point in space and time.

@@

\noindent In probability and statistics, random variables are defined by specifying the **distribution** the variable is assumed to follow. Since the variable is random, it can take multiple values, each with different probabilities. Commonly known distributions include the normal (i.e., Gaussian) distribution, the Student's T distribution, and the Bernoulli distribution.

The set or range of values which have non-zero probabilities in a distribution are referred to as the distribution's **support**. The support of the normal and T distributions are all real numbers (i.e., from $-\infty$ to $\infty$), while the Bernoulli distribution only has support over values 0 and 1.

There are many distributions we can assign to $R$, depending on how we quantify it. If $R$ is the number of calories an animal is able to access from food in a given location, we can let $R$ follow any distribution with support over all real numbers between 0 and infinity, which we can indicate with $R \in (0, \infty)$.^[We use round parentheses rather than square brackets (i.e., $[0, \infty]$) because neither 0 nor infinity are included in the distribution's support set.] In this case, we can let $R$ follow a Gamma or log-normal distribution. If $R$ was a discrete variable, such as the number of prey in a location during a period of time, we can use a Poisson or negative binomial distribution. Alternatively, if we measure $R$ using the Normalized Difference Vegetation Index [**NDVI**, see @pettorelli_normalized_2011], we should use a distribution with support over the interval $[-1, 1]$, since NDVI can only take on values between -1 and 1 (extremes included). The next two sections introduce two parameters that are often used to visualize, compare, and define distributions: the expectation and the variance of a random variable.

@@

\noindent Since the exact value of $R$ at a given time and location is unknown, comparing the magnitude in $R$ between two locations or time periods requires a quantitative measure of what value we believe $R$ will take, on average. The **expectation** of a random variable (i.e. its mean or average, indicated here with the symbol $\e$) is the value one can *expect* the random variable to take. Thus, we use $\e(R) = \mu$ to indicate the expectation of the random variable $R$. The mean may not be within the distribution's support (e.g., a non-integer number of prey per day), but $\e(R)$ is still useful for estimating the average long-term resource abundance. When the mean changes over time, as in the case changes in $R$ in highly seasonal regions, we explicitly indicate that $\e(R) = \mu$ changes over time by writing the expectation of $R$ as a function of time, $t$: $\e(R) = \mu(t)$. We indicate the estimated average resource abundance by adding a caret on the symbol: $\hat \mu(t)$.

The hypothesis that spatial needs decrease with resource abundance, $R$, is commonly accepted and well supported, but many studies assume a linear relationship [e.g., @harestad_home_1979; @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022; @bradsworth_using_2022; @mcclintic_effects_2014]. This is problematic because it can lead to questionable estimates and problematic biases, such as negative or excessively small home ranges [e.g., figure 2 in the work of @bista_effect_2022]. Conceptually, the relationship between spatial needs and resource abundance must be nonlinear, since there are limits to how much space an animal is able to explore in its lifetime and the minimum amount of space it requires to survive (which is necessarily greater than zero). As $\e(R)$ increases, we expect animals' spatial requirements to decrease nonlinearly until they reach the smallest amount of space required to survive [see @relyea_home_2000; @bista_effect_2022, although the latter uses models which assume linear relationships]. The spatial requirements initially decrease rapidly since small additive increases (e.g., from 1 to 2) are proportionally large (i.e., 2 is twice 1). In contrast, at higher values of $\e(R)$, the same additive changes (e.g. from 50 to 51) are proportionally small (i.e., 51 is only 2% larger than 50).

@@

\noindent A random variable's **variance** is a measure of its unpredictability or variability. We use  $\var(R) = \sigma^2$ to indicate the variance in $R$, after accounting for changes in $\mu(t)$, and $\widehat{\sigma^2}(t)$ to indicate its estimate conditional on time. For instance, while one may consider $R$ to vary within years, we define $\var(R)$ as the variation in $R$ after accounting for the fact that $R$ is lower during some seasons and higher in others. Additionally, we distinguish between two different sources of variation. We refer to known differences or expected spatiotemporal changes as **heterogeneity**, while we use the term **stochasticity** for unpredictable and unexpected variation. For example, the location of (high-yield) berry bushes in a valley and the time at which they produce fruit may be heterogeneous, but predictable once an animal visits the valley. In contrast, whether next year will be a good or bad year for berries is stochastic.

@@

\noindent In this paper, we simulate $R$ using a Gamma distribution with time-dependent mean $\mu(t)$ and time-dependent variance $\sigma^2(t)$ (which we write as $R \sim \Gamma(\mu(t), \sigma^2(t))$). Although Gamma distributions are more often parameterized using parameters shape and scale ($k, \theta$) or shape and rate ($\alpha, \beta = 1/\theta$), we use $\mu = k \theta$ and $\sigma^2 = k \theta^2$ to facilitate visualizing the examples. However, note that $\mu$ and $\sigma^2$ are not independent because the variance depends strongly on the mean. As the mean approaches zero (from the positive side, which we write as $\mu \rightarrow0^+$) the variance also does: $\mu \rightarrow 0^+ \iff k \theta \rightarrow 0^+ \iff k \theta^2 = 0^+ \iff \sigma^2 = 0^+$.

@@

\noindent In this paper's empirical example, we estimate resource abundance using NDVI [for more information, see the relevant section later in the paper and @pettorelli_normalized_2011]. However, since there is no commonly used distribution with a support over the interval $[-1, 1]$ we scale NDVI to the interval $[0, 1]$ to use a Beta distribution by applying the linear transformation

\begin{equation} \label{ndvi-to-r}
R = \frac{\nu + 1}{2},
\end{equation}

where $\nu$ indicates the NDVI value. Note that only **linear transformations** (i.e. only addition, subtraction, multiplication, and division) should be applied to random variables to avoid biases while estimating the distribution and its parameters [@denny_fallacy_2017; @jensen_sur_1906]. In this case, we can define $R$ as following a Beta distribution with mean and variance that depend on time using the notation $B (\mu(t), \sigma^2(t))$. We use this parameterization here for ease of explanation, but note that Beta distributions are generally parameterized using the shape parameters $\alpha$ and $\beta$ such that the mean is

\begin{equation} \label{beta-mean}
\mu = \frac{\alpha}{\alpha + \beta}
\end{equation}

@@

\begin{equation} \label{beta-var}
\sigma^2 = \frac{\alpha\beta}{(\alpha+\beta)^2(\alpha+\beta+1)}.
\end{equation}

\noindent Failing to model or simulate resource abundance appropriately can fail to produce robust, sensible, and accurate results.

## Animal movement as a stochastic process

@@

\noindent @nathan_movement_2008 generalized animal movement from location $\vec l_t$ to location $\vec l_{t+1}$ using the model

\begin{equation} \label{nathan-movement}
\vec l_{t+1} = F(\Omega, \Phi, \vec w_t, \vec l_t, \vec \xi_t),
\end{equation}

\noindent such that an animal's position as time $t+1$ depends on its motion capacity ($\Omega$), navigation capacity ($\Phi$), internal state ($\vec w_t$) and current location ($\vec l_t$), as well as how each of the parameters may change in response to the current environmental factors ($\vec \xi_t$). This model is conceptually useful, as it can form the basis for many models, experiments, and hypotheses, but fitting such a model would be very difficult in practice (if not impossible). The parameters $\Omega, \Phi, \vec w_t$ and $\vec \xi_t$ cannot be quantified directly, and although it may be possible to monitor proxies for the parameters (e.g., body size, memory, cortisol levels), doing so with sufficient frequency and accuracy would not be simple. Thus, we can instead focus on animals' behavior and spatial needs as a response to overall resource abundance and unpredictability.

@@

We start by providing examples and considerations on how one may model $R$, animals' spatial needs, and the effects of $\e(R)$ and $\var(R)$ on animals' spatial needs. Next we provide an empirical example using tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado and NDVI as a proxy for $R$. Finally, we offer suggestions on how this approach can be used to inform conservation-related decisions, including assessing habitat quality and estimating animals spatial needs under different scenarios.

@@

# MISC

>>>>

---
title: \large How resource abundance and stochasticity affect animals' spatial needs
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: [braes, biol]
  - name: Simon Wood
    institute: uofe
  # - name: Adam T. Ford
  #   institute: biol
  #   institute: braes
  # - name: Jacob R. Goheen
  #   institute: wyom
  - name: E. Patrícia Medici
    institute: [tapir, iucn, escas]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - uofe: School of Mathematics, James Clerk Maxwell Building, University of Edinburgh, Edinburgh, Unted Kingdom.
  # - wyom: Department of Zoology and Physiology, University of Wyoming, Laramie, Wyoming, United States of America.
  - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
  - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
  - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'hr-resource-stoch.bib'
csl: 'the-american-naturalist.csl'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{wrapfig} # for wrapping text around figures
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{afterpage} # for \afterpage{\<command>}, e.g. \afterpage{\clearpage}
subparagraph: true # needed for \usepackage[small]{titlesec}
linenumbers: true
output:
  bookdown::pdf_document2:
    latex_engine: xelatex # @reed_frequency_2003 has unicode character U+0308
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: false
editor_options: 
  chunk_output_type: console
---

@@

@@

@@

\noindent Animals' spatial needs are thought to be tightly linked to the amount of resources within their habitats, such that animals living in productive habitats generally require less space than those in resource-poor habitats. This hypothesis has widespread empirical support, but existing studies have focused primarily on responses to the *mean* amount of resources, while responses to the variance around the mean are still largely unknown. This is not a trivial oversight. Animals adjust to variable environmental conditions, so failing to account for the effects of resource unpredictability can result in a limited understanding of animals' spatial needs. In this study, we first review the literature of studies that estimate animals' spatial needs as a function of resource abundance. We then leverage this information to provide a unifying framework and hypotheses for the effect of mean *and* variance in resources on animals' space use. Next, we use simulated movement data to demonstrate how the combined effects of mean and variance in resource abundance interact to shape predictable patterns in animal space use. Finally, we use real-world tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado <!--an impala (*Aepyceros melampus*), and multiple moose (*Alces alces*)--> to show how this framework can be applied to better understand the movement ecology of free-ranging animals. Results from the simulations and empirical examples are presented using a fully transparent approach that allows researchers to apply the framework to their own data and inform area-based conservation efforts.

@@

\noindent The amount of resources an organism is able to access is a strong determinant of its odds of survival and reproduction. Resource limitations can cause individuals to experience a negative energetic balance, which can then result in lower fitness [@hou_cold_2020; @le_bot_fishery_2019], changes to community structure [@haney_tipping_2018; @burson_competition_2018; @riotte-lambert_environmental_2020; @ghislandi_resource_2018], altered physiology [@dai_pra_ground_2022; @rocha_life_2021; @le_bot_fishery_2019; @wessling_seasonal_2018], lower chance of reproduction [@stefanescu_timing_2021; @schmidt_interplay_2020; @le_bot_fishery_2019; @douglas_relative_2014], and even death [@foley_severe_2008; @berger_climate_2018]. Thus, many organisms will adapt their behaviors in response to changes in local resource abundance to ensure their needs are met. Some species may respond to fluctuations in resource abundance by, if possible, switching to other food sources [@steinmetz_effects_2021; @le_bot_fishery_2019], reducing energetic costs by reducing body heat [@schmidt_interplay_2020], or entering hibernation or torpor [@fjelldal_nightly_2021; @mohr_cellular_2020; @boyles_optimal_2020], but movement represents one of the most readily available traits that species can adjust, whether this be by modifying their home range [@bista_effect_2022; @bradsworth_using_2022; @lucherini_habitat_1996; @relyea_home_2000; @yu_artificial_2022; @arechavala-lopez_common_2019], migrating [@middleton_green-wave_2018; @geremia_migrating_2019], dispersing [framework: @southwood_habitat_1977; amphibians: @cayuela_determinants_2020; birds: @pretorius_movement_2020; @wheat_migrate_2017; mammals: @singh_migration_2012], or moving nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019].

The relationship between animal movement and resource abundance has been of interest to biologists for over half a century. @burt_territoriality_1943 considered the search for food as the primary driver for animal movement within an animal's home range. Three decades after, @southwood_habitat_1977 suggested that change in resource abundance is a strong determinant of how animals decide where to live and when to reproduce. Two years later, @harestad_home_1979 proposed that the simplest relationship between resource abundance and an animal's home-range size is

\begin{equation}
H = C / R,
(\#eq:harestad-eq)
\end{equation}

\noindent where $H$ is the animal's **home-range size**, $C$ is the animal's resource consumption, and $R$ is the **resources** an animal can access. However, a species' spatial requirements depend strongly on its metabolic rate, diet, and body weight [@harvey_primate_1981; @lindstedt_home_1986; @reiss_scaling_1988; @harvey_primate_1981; @gittleman_carnivore_1982; @boratynski_energetic_2020; @noonan_effects_2020].

While many researchers have demonstrated that animals adapt heir home ranges in response to resources abundance, few studies build upon others' previous work. Instead, results are reported as independent, *de-novo* findings. Consequently, we currently lack a unifying framework for quantifying the effects of resource abundance on animals' spatial needs. In addition, while much work has been done on estimating animals' responses to average resource abundance, there is little to no information on how they respond to unpredictable changes in resources.

In this paper, we refer to a location's average amount of resources as **resource abundance**, while use the phrase **resource stochasticity** to indicate the variability in resource abundance due to unpredictable causes. We argue that, on its own, a habitat's long-term resource abundance is not sufficient to assess the habitat's quality. For instance, a grassland with relatively low but constant forage availability will clearly require drastically different behaviors and adaptations from a desert location with equally scarce forage but rare, sudden, and strong pulses of resources. In the first scenario, an animal may require a large but constant home range as it moves between locations in search of food, while an animal in the second scenario may switch between dispersal as it searches for high-resource patches and short-term range residency until the local resources are depleted. Although it may be possible for both habitats to have the same long-term average resource abundance, the differences in resource unpredictability result in substantially different movement strategies being selected for. Although it is generally expected that resource unpredictability will decrease animals' fitness and a landscape's energetic balance [@chevin_adaptation_2010], there is little empirical evidence to support this hypothesis [but see: @herfindal_prey_2005; @nilsen_can_2005; @rizzuto_forage_2021].

Here, we illustrate how animals' spatial needs depend on the abundance and unpredictability of resources. First, we set the theoretical background necessary for the successive sections by introducing concepts and notations without assuming any previous knowledge of statistics or movement ecology. Next, we provide a review of the effects of resource *abundance* on animals' spatial needs while suggesting a unifying hypothesis. Afterwards, we provide a review of the effects of resource *stochasticity* on animals spatial needs while suggesting a second unifying hypothesis. Subsequently, we test the two hypotheses using quantitative, simulated responses to changes in resource abundance and unpredictability. Finally, we demonstrate how this framework can be used in practice to describe the movement ecology of a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado [@medici_movement_2022]. Results from the simulations and empirical example are presented using a fully transparent approach that allows researchers to replicate the work and apply the methods to their own tracking data.

@@

<!--\noindent In this section, we presents the foundational concepts and terms used in this manuscript. We introduce each concept without assuming a statistical or quantitative background to maximize accessibility and minimize misunderstandings and misinterpretations.-->

@@

\noindent In statistics, **random variables** indicate random (i.e., unknown) quantities and are indicated with capital letters (e.g., $R$). Known values, such as realizations of random variables (i.e., known observations or instances), are indicated with lower-case letters (e.g., $r$). Using this notation, we can write the statement "the probability of random variable $R$ taking the value $r$" as $P(R = r)$. Resource abundance is often unpredictable (and hard to quantify), since it depends on a variety of factors which cannot be accounted for easily, including climate, weather, competitive pressure, and individual- and species-specific energetics. Thus, we can let the random variable $R$ indicate the resource abundance at a given point in space and time.

@@

\noindent Random variables are defined by specifying the **distribution** the variable is assumed to follow (e.g., Gaussian, Gamma, Poisson, Beroulli). Since the variable is random, it can take multiple possible values, each with different probabilities. The set or range of values which have non-zero probabilities in a distribution are referred to as the distribution's **support**.

There are many distributions we can assign to $R$, depending on how we quantify it. If $R$ is the number of calories an animal is able to access from food in a given location, we can let $R$ follow a distribution with support over all real numbers between 0 and infinity, which we can indicate with $R \in (0, \infty)$,^[We use round parentheses rather than square brackets (i.e., $[0, \infty]$) because neither 0 nor infinity are included in the distribution's support set.] such as a Gamma or log-normal distribution. If $R$ is a discrete variable, such as the number of prey in a location during a period of time, we can use a Poisson or negative binomial distribution. Alternatively, if we measure $R$ using the Normalized Difference Vegetation Index [**NDVI**, see @pettorelli_normalized_2011], we should use a distribution with support over the interval $[-1, 1]$, since NDVI can only take on values between -1 and 1 (extremes included).

@@

\noindent Since the exact value of $R$ at a given time and location is unknown, comparing the magnitude in $R$ between two locations or time periods requires a quantitative measure of what value we believe $R$ will take, on average. The **expectation** of a random variable is the value one can *expect* the random variable to take, on average, in the long term. Thus, we can let $\e(R)$ indicate the expectation of the random variable $R$. When the mean changes over time, such as in strogly seasonal regions, we explicitly indicate that $\e(R)$ changes over time by writing $\e(R)$ as a function of time, $t$: $\mu(t)$. We indicate the *estimated* average resource abundance over time by adding a caret on the symbol: $\hat \mu(t)$.

The hypothesis that spatial needs decrease with resource abundance, $\e(R)$, is commonly accepted and well supported, but many studies assume a linear relationship [e.g., @harestad_home_1979; @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022; @bradsworth_using_2022; @mcclintic_effects_2014]. This is problematic because a linear relationship can lead to questionable estimates and problematic biases, such as negative or excessively small home ranges [e.g., figure 2 in the work of @bista_effect_2022]. Conceptually, the relationship between spatial needs and $\e(R)$ must be nonlinear, since there are limits to how much space an animal is able to explore in its lifetime as well as the minimum amount of space it requires to survive (which is necessarily greater than zero). As $\e(R)$ increases, we expect animals' spatial requirements to decrease nonlinearly until they reach the smallest amount of space required to survive [e.g., @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005]. <!--The spatial requirements initially decrease rapidly since small additive increases (e.g., from 1 to 2) are proportionally large (i.e., 2 is twice 1). In contrast, at higher values of $\e(R)$, the same additive changes (e.g. from 50 to 51) are proportionally small (i.e., 51 is only 2% larger than 50), so spatial needs decrease more slowly.-->

@@

\noindent In viewing resources as a random variable, we not only obtain a formal framework for describing $\e(R)$, but also the spread around $\e(R)$. A random variable's **variance** is a measure of its unpredictability (i.e., variability). We use the notation $\var(R)$ to indicate the variance in $R$ after accounting for changes in $\mu(t)$ and $\widehat{\sigma^2}(t)$ to indicate its estimate over time. Consequently, while one may consider $R$ to vary within years, we define $\var(R)$ as the variation in $R$ after accounting for the fact that $R$ is (predictably) lower during some seasons and higher in others. Additionally, we distinguish between two different sources of variation: We refer to known or expected spatiotemporal differences or changes as **heterogeneity**, which we consider as a change in $\mu(t)$. In contrast, we use the term **stochasticity** for unpredictable and unexpected variation (i.e., due to $\sigma^2(t)$. For example, the location of (high-yield) berry bushes in a valley and the time at which they produce fruit may be heterogeneous, but it becomes predictable once a bear visits all patches in the valley. In contrast, whether next year will be a good or bad year for berries is stochastic because it depends on factors the bear cannot predict.

@@

\noindent In this paper, we simulate $R$ using a Gamma distribution with time-dependent mean $\mu(t)$ and time-dependent variance $\sigma^2(t)$, which we write as $R \sim \Gamma\big(\mu(t), \sigma^2(t)\big)$. Although Gamma distributions are more often parameterized using parameters shape and scale ($k, \theta$) or shape and rate ($\alpha, \beta = 1/\theta$), we use $\e(R) = k \theta$ and $\var(R) = k \theta^2$ to facilitate visualizing the simulations. However, note that $\e(R)$ and $\var(R)$ are not independent because the variance depends strongly on the mean (and vice-versa): As the mean approaches zero (from the positive side, which we write as $\e(R) \rightarrow0^+$) the variance also does: $\e(R) \rightarrow 0^+ \iff k \theta \rightarrow 0^+ \iff k \theta^2 = 0^+ \iff \var(R) = 0^+$. This assumption also holds biologically: In the example from above, a mean of zero kg of berries within the bear's habitat during winter implies the variance in berries is also zero.

@@

\noindent In this paper's empirical example, we estimate resource abundance using NDVI [for more information, see the relevant section later in the paper and @pettorelli_normalized_2011]. However, since there is no commonly used distribution with a support over the interval $[-1, 1]$ we scale NDVI to the interval $[0, 1]$ to use a Beta distribution by applying the **linear transformation**

\begin{equation}
\nu_{[0, 1]} = \frac{\nu + 1}{2},
(\#eq:ndvi-to-01)
\end{equation}

\noindent where $\nu$ indicates the NDVI value [see @denny_fallacy_2017 for more information on transforming response data]. <!--Note that only **linear transformations** (i.e. only addition, subtraction, multiplication, and division) should be applied to random variables to avoid biases while estimating the distribution and its parameters [@denny_fallacy_2017; @jensen_sur_1906; @ohara_not_2010].--> In this case, we can define $R$ as following a Beta distribution with mean and variance that depend on time using the notation $B (\mu(t), \sigma^2(t))$. We use this parameterization here for ease of explanation, but note that Beta distributions are generally parameterized using the shape parameters $\alpha$ and $\beta$ such that the mean is

\begin{equation}
\e(R) = \frac{\alpha}{\alpha + \beta}
(\#eq:beta-mean)
\end{equation}

@@

\begin{equation}
\var(R) = \frac{\alpha\beta}{(\alpha+\beta)^2(\alpha+\beta+1)}.
(\#eq:beta-var)
\end{equation}

\noindent We can easily convert $\mu(t)$ and $\sigma^2(t)$ back to mean and variance in NDVI using the inverse of equation \@ref(eq:ndvi-to-01):

\begin{equation}
\nu = 2 \nu_{[0, 1]} - 1.
(\#eq:01-to-ndvi)
\end{equation}

\noindent Consequently, we have

\begin{equation}
\e(\nu|t) = 2 \mu(t) - 1.
(\#eq:mu-to-e-ndvi)
\end{equation}

\noindent and

\begin{equation}
\var(\nu|t) = 2^2 \sigma^2(t) + 0 = 4 \sigma^2(t).
(\#eq:sigma2-to-v-ndvi)
\end{equation}

\noindent Particular attention should be be given when deciding what distribution to use and how to estimate means and variances in $R$. Improper models and simulations of resource abundance can fail to produce robust, sensible, and accurate estimates of $R$.

<!--
## Animal movement as a stochastic process

@@

\noindent @nathan_movement_2008 suggested we can generalize animal movement from location $\vec l_t$ to location $\vec l_{t+1}$ using the model

\begin{equation}
\vec l_{t+1} = F(\Omega, \Phi, \vec w_t, \vec l_t, \vec \xi_t),
(\#eq:nathan-movement)
\end{equation}

\noindent such that an animal's position as time $t+1$ depends on its motion capacity ($\Omega$), navigation capacity ($\Phi$), internal state ($\vec w_t$) and current location ($\vec l_t$), as well as how each of the parameters may change in response to the current environmental factors ($\vec \xi_t$). This model is conceptually useful, as it can form the basis for many models, experiments, and hypotheses, but fitting such a model would be very difficult in practice (if not impossible). The parameters $\Omega, \Phi, \vec w_t$ and $\vec \xi_t$ cannot be quantified directly, and although it may be possible to monitor proxies for the parameters (e.g., limb length, body size, site fidelity, cortisol levels, reproductive stage), doing so with sufficient frequency and accuracy would not be simple. Thus, we can instead focus on animals' movement and spatial needs as a response to overall resource abundance and unpredictability while accepting other drivers of movement as causes of stochastic behaviors.

@@

We start by providing examples and considerations on how one may model $R$, animals' spatial needs, and the effects of $\e(R)$ and $\var(R)$ on animals' spatial needs. Next, we provide an empirical example using tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado and NDVI as a proxy for $R$. Finally, we offer suggestions on how this approach can be used to inform conservation-related decisions, including assessing habitat quality and estimating animals spatial needs under different scenarios.

@@
